Towards a miniKanren with fair search strategies

### 2019-05-22 Paper/paper.tex
Someone should use the tex macros that have already been written and get rid of looking at these extra long objects that require a steady hand.  They are all over the place, for example \DFSi{}.  I am up to 2.2 on perhaps my final pass over the paper before it is submitted.

  We describe fairness levels in disjunction and conjunction
  implementations.  Specifically, a disjunction implementation can be
  fair, almost-fair, or unfair. And a conjunction implementation can
  be fair or unfair.  We compare the fairness level of four search
  strategies: the standard miniKanren interleaving depth-first search (\DFSi),
  the balanced interleaving depth-first search (\DFSbi), the fair depth-first
  search (\DFSf), and the standard breadth-first search (\BFSser).  
  \DFSbi{} and \DFSf{} are new. And we present a new implementation of 
  \BFSser. Using a quantitative evaluation, we argue that
  \DFSbi{} and \DFSf{} are competitive alternatives to \DFSi, 
  and that \BFSopt{} is more efficient than the standard \BFSser.

@@

\NOTE[DPF]{We should reference the earliest paper (See Will Byrd vita)

@@

@@

The miniKanren in TR2 implements interleaving DFS (\DFSi), the cause of this 
unexpected result. With this search strategy, each clause takes half 
of its received computational resources and pass the other half to its 
following clauses, except for the last clause that takes all resources it 
receives. In the example above, the \texttt{a} clause takes half of all 
resourses. And the \texttt{b} clause takes a quarter. Thus \texttt{c} and 
\texttt{d} barely contribute to the result.

@@

There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves
for all purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When the type checker is used to generate simple and shallow programs,
the clauses of constructors should be put in the front of \conde.
When performing proof searches for complicated programs, the clauses of 
eliminator should take the focus. With \DFSi, these two uses cannot be 
efficient at the same time. In fact, to make one use efficient, the other one 
must be drastically slow.

The specification that every clause in the same \conde{} is given equal 
``search priority'' is called fair \disj{}. And search strategies with 
almost-fair \disj{} give every clause similar priority. 
Fair \conj{}, a related concept, is more subtle. We defer it to the next 
section.

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we would like to give a short review about 
state, search space, and goal, because fairness is defined in terms of them. 
A \emph{state} is a collection of constraints. Every answer corresponds to a 
state. A \emph{search space} is a collection of states. And a \emph{goal} is a 
function from a state to a search space.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the query result. \DFSi, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters  (e.g. \texttt{c} and \texttt{d}) are
rarely seen. The situation would be exacerbated if \conde{} contains more 
clauses.

@@

Under the hood, the \conde{} here is allocating computational resources to 
four trivially different search space. The unfair \disj{} in 
\DFSi{} allocates many more resources to the first search space. On the 
contrary, fair \disj{} would allocate resources evenly to each search space.

@@

We end this subsection with precise definitions of all levels of 
\disj{} fairness. Our definition of \emph{fair} \disj{} is slightly generalize 
from the one by \citet{seres1999algebra}. Their definition is for binary 
disjunction. We generalize it to a multi-arity one.

\begin{defn}[fair \disj{}]
A \disj{} is fair if and only if it allocates computational resources evenly to 
search spaces produced by goals in the same disjunction (i.e. clauses in 
the same \conde).
\end{defn}

@@

\begin{defn}[unfair \disj{}]
A \disj{} is unfair if and only if it is not even almost-fair.
\end{defn}

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. Search strategies with unfair \conj{} (e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde{} contains more clauses.

\NOTE[LKC]{Should I add a footnote?

``
Although 
DFS$_\textrm{i}$'s \disj{} is unfair in general, it is fair when there is no 
call to relational definition in \conde{} clauses (including this case).
''}

>>>>

  We describe fairness levels in disjunction and conjunction
  implementations.  Specifically, a disjunction implementation can be
  fair, almost-fair, or unfair. And a conjunction implementation can
  be fair or unfair.  We compare the fairness level of four search
  strategies: the standard miniKanren interleaving depth-first search (\DFSi),
  the balanced interleaving depth-first search (\DFSbi), the fair depth-first
  search (\DFSf), and the standard breadth-first search (\BFSser).  
  \DFSbi{} and \DFSf{} are new. And we present a new, more efficient
  and simpler implementation of \BFSser. Using quantitative evaluation, 
  we argue that
  \DFSbi{} and \DFSf{} are competitive alternatives to \DFSi, 
  and that \BFSopt{} is more efficient than the standard \BFSser.

@@

@@

\NOTE[DPF]{We should reference the earliest paper (See Will Byrd vita)}

@@

The miniKanren in TRS2 implements interleaving DFS (\DFSi), the cause of this 
unexpected result. With this search strategy, each clause takes half 
of its received computational resources and passes the other half to its 
following clauses, except for the last clause that takes all resources it 
receives. In the example above, the \texttt{a} clause takes half of all 
resourses. And the \texttt{b} clause takes a quarter. Thus \texttt{c} and 
\texttt{d} barely contribute to the result.

@@

There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves all purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When the type checker is used to generate simple and shallow programs,
the clauses for constructors should be put at the top of the
\conde{} clauses.
When performing proof searches for complicated programs, the clauses for 
eliminators should be put at the top of \conde{} clauses. With \DFSi, these two uses cannot be 
efficient at the same time. In fact, to make one use efficient, the other one 
must be more sluggish.

The specification that every clause in the same \conde{} is given equal 
``search priority'' is called fair \disj{}. And search strategies with 
almost-fair \disj{} give every clause similar priority. 
Fair \conj{}, a related concept, is more subtle. The discussion of fair
\conj{} is covered in the next section.

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we give a short review of the terms:
\emph{state}, search \emph{space}, \emph{goal}.
A \emph{state} is a collection of constraints. (Here, we restrict 
constraints to unification constraints. Every answer corresponds to a 
state. A space is a collection of states. And a \emph{goal} is a 
function from a state to a space.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the query result. \DFSi, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters  (e.g. \texttt{c} and \texttt{d}) are
rarely seen. The situation would be exacerbated if \conde{} would have
contained more clauses.

@@

Under the hood, the \conde{} here is allocating computational resources to 
four trivially different search spaces. The unfair \disj{} in 
\DFSi{} allocates many more resources to the first search space. On the 
contrary, fair \disj{} would allocate resources evenly to each search space.

@@

We end this subsection with precise definitions of all levels of 
\disj{} fairness. Our definition of \emph{fair} \disj{} is slightly 
more general
than the one in \citet{seres1999algebra}. Their definition is only
for binary disjunction. We generalize it to a multi-arity one.

\begin{defn}[fair \disj{}]
A \disj{} is fair if and only if it allocates computational resources evenly to 
search spaces produced by goals in the same disjunction 
(i.e., clauses in the same \conde).
\end{defn}

@@

\begin{defn}[unfair \disj{}]
A \disj{} is unfair if and only if it is not almost-fair.
\end{defn}

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. Search strategies with unfair \conj{} (e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde{} were to contain more clauses.

\NOTE[LKC]{Should I add a footnote?}

``
Although 
DFS$_\textrm{i}$'s \disj{} is unfair in general, it is fair when there is no 
call to a relational definition 
in \conde{} clauses (including this case).
